
%=======================================================================
% riscv-interrupts.tex
%-----------------------------------------------------------------------

\documentclass[twoside,11pt]{article}

% Fix copy/pasting of ligatures in Acrobat
\input{glyphtounicode.tex}
\pdfgentounicode=1 %

\input{preamble}

\newcommand{\DuoPLICRev}{0.3.3-draft}

\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}

%=======================================================================

\begin{document}

%=======================================================================

\title{%
  \vspace{30mm}
  {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\
  {\large Document Version \DuoPLICRev}
}

\author{%
  Editor: John Hauser \\
  {\tt jh.riscv\mbox{}@jhauser.us} \\
  \\
  \today
}

\date{}
\maketitle

\thispagestyle{empty}

\vfill

\textbf{%
Warning!
This document is only a draft and might never be
accepted by the {\RISCV} International Association.
If accepted, the Duo-PLIC specification here is
liable to change before becoming a ratified standard.%
}

\vfill

Contributors to all versions of this specification in alphabetical
order (please contact the editor to suggest corrections):
John Hauser.

This document is released under a Creative Commons Attribution 4.0
International License.

\vfill

\clearpage

%=======================================================================

The Advanced PLIC of the {\RISCV} Advanced Interrupt Architecture
is not backward compatible with the original {\RISCV} PLIC.
To ease the transition of software from the original PLIC to the
APLIC, a {\RISCV} system may have a \emph{\mbox{Duo-PLIC\/}}
that software can configure to act as either form of PLIC.
A \mbox{Duo-PLIC} is an extension of an APLIC, having all the
usual memory-mapped interfaces of an APLIC, plus, at some other
memory addresses, the memory-mapped interface of an original {\RISCV}
PLIC too.
When the \mbox{Duo-PLIC} has \emph{compatibility mode} enabled,
interrupts from the \mbox{Duo-PLIC} to harts are controlled exclusively
by the original PLIC side of the \mbox{Duo-PLIC};
and when compatibility mode is disabled, interrupts to harts
are controlled exclusively by the APLIC side of the
\mbox{Duo-PLIC}.
A \mbox{Duo-PLIC} comes out of reset with compatibility mode enabled,
so older software with no knowledge of an APLIC can see the
\mbox{Duo-PLIC} as an original PLIC.

When compatibility mode is enabled so a \mbox{Duo-PLIC} acts as an
original PLIC, the \mbox{Duo-PLIC} supports only the direct delivery of
interrupts to harts, not the forwarding of interrupts by MSIs.
When compatibility mode is disabled and the \mbox{Duo-PLIC} acts as
an APLIC, direct delivery mode is supported for all interrupt
domains, and MSI delivery mode may optionally be supported.
Hence, for each APLIC interrupt domain, field DM
(Delivery Mode) of the \z{domaincfg} register is either read-only zero
or writable, but is not read-only one.

In the memory-mapped control region for the APLIC root
interrupt domain (only), a \mbox{Duo-PLIC} adds to the \z{domaincfg}
register a CM (Compatibility Mode) field in bit~6.
The complete format of the root interrupt domain's \z{domaincfg} is
thus:
\begin{displayLinesTable}[l@{\quad}l]
bits 31:24 & read-only \z{0x80} \\
bit 8      & IE \\
bit 7      & read-only 0 \\
bit 6      & CM \\
bit 2      & DM (\WARL) \\
bit 0      & BE (\WARL) \\
\end{displayLinesTable}
All other register bits are reserved and read as zeros.

Reset initializes the APLIC root domain's \z{domaincfg}
register with CM =~1 and with all other writable bits zeros.

When CM~=~1 (compatibility mode enabled):
\begin{itemize}

\item
For each \emph{context} implemented by the original PLIC side of the
\mbox{Duo-PLIC}, the output interrupt for that context becomes the
interrupt signal delivered from the \mbox{Duo-PLIC} to the specific
hart and privilege level specified by the implementation.
Other than the CM field of the APLIC root domain's
\z{domaincfg} register, the state of the APLIC side of the
\mbox{Duo-PLIC} is irrelevant to the interrupts delivered to harts.

\item
In the memory-mapped control region for each APLIC interrupt
domain:
\begin{itemize}

\item
Fields IE and DM of the \z{domaincfg} register are read-only zeros.

\item
Field BE of the \z{domaincfg} register has its usual function for the
interrupt domain but has no effect on accesses to the registers of the
original PLIC side of the \mbox{Duo-PLIC}.

\item
If they exist, the registers that configure MSI addresses
(\z{mmsiaddrcfg}, \z{mmsiaddrcfgh}, \z{smsiaddrcfg}, and
\z{smsiaddrcfgh}) remain functional.

\item
All other registers are read-only zeros.
Writes to the registers are ignored.

\end{itemize}

\end{itemize}

When CM~=~0 (compatibility mode disabled):
\begin{itemize}

\item
Interrupts to harts are determined by
the APLIC side of the \mbox{Duo-PLIC}.
The state of the original PLIC side of the \mbox{Duo-PLIC} is
irrelevant.

\item
In the memory-mapped control region for the original PLIC side of the
\mbox{Duo-PLIC}, all registers are read-only zeros.
Writes to the registers are ignored.

\end{itemize}

If the \mbox{Duo-PLIC}'s compatibility mode is changed from disabled to
enabled, the registers of the original PLIC side of the \mbox{Duo-PLIC}
obtain valid and consistent but otherwise {\unspecified} values.

If the \mbox{Duo-PLIC}'s compatibility mode is changed from
enabled to disabled, the registers of the APLIC side of
the \mbox{Duo-PLIC} obtain valid and consistent but otherwise
{\unspecified} values, except for the \z{domaincfg} registers
and, if they exist, the registers that configure MSI addresses
(\z{mmsiaddrcfg}, \z{mmsiaddrcfgh}, \z{smsiaddrcfg}, and
\z{smsiaddrcfgh}), all of which retain their previous values.
(The root domain's \z{domaincfg} gets the value written when changing
field~CM, of course.)

The endianness of the memory-mapped registers of the original PLIC side
of the \mbox{Duo-PLIC} is not affected by the BE field of any
APLIC \z{domaincfg} register.
For most, maybe all versions of an original PLIC, the registers are
always little-endian.

The number of interrupt priority levels supported by the original PLIC
side of a \mbox{Duo-PLIC} is the same as that of the APLIC side
when in direct delivery mode.
Hence, the number of integer bits to store a priority number, IPRIOLEN,
is a constant for a \mbox{Duo-PLIC} (in the range 1 to 8 as usual),
independent of whether compatibility mode is enabled or disabled.

However, an original PLIC encodes priorities in the opposite order
from an IMSIC or APLIC, with number 1 being interpreted as the
lowest priority and larger numbers as higher priorities.
When a \mbox{Duo-PLIC} is acting as an original PLIC
and reports to a hart an interrupt priority number,
the number reported is negated to \emph{normalize} it to
the convention of the Advanced Interrupt Architecture.
If the original PLIC's priority number for an interrupt is~$p$,
the normalized priority number communicated to the target hart is
${\mbox{2}^\textrm{IPRIOLEN}-p}$.

\begin{commentary}
Negation of original PLIC priority numbers to normalize them matters
only when a hart supports configurability of the priorities of major
interrupts, an option added by the Advanced Interrupt Architecture.
Software that is oblivious to the Advanced Interrupt Architecture
should have no reason to see priority numbers being negated.
\end{commentary}

%=======================================================================

\end{document}

